Circular Regression Trees and Forests with an Application to Probabilistic Wind Direction Forecasting

### 2019-12-09 pkg/circtree/inst/wind_paper/draft/draft.Rnw
Include main issues of Georg's comments
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2921 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 91c5c1244c11c460873fa3a9b5e4cce158213c61

\Abstract{
Despite the occurrence of circular data in a wide range of scientific fields,
the existing methodology for probabilistically forecasting circular data is
highly limited. Most of the existing methods are built on the framework of
generalized linear and additive models, which can incorporate difficulties in
both the optimization and the interpretation. As a very intuitive alternative
to additive circular regression models, we extend previous ideas of circular
trees and introduce probabilistic circular regression trees and forests
employing the von Mises distribution. The resulting tree-based models simplify
the estimation process by separating the data into more homogeneous subgroups
and fitting a simple intercept-only model for each resulting subset. By
incorporating the framework of distributional modeling, splits in all
distribution parameters are considered and a fully specified distribution is
provided for each terminal node. The proposed circular regression trees are
easy to interpret, can detect non-additive effects, and select covariates and
their interactions automatically. By combining an ensemble of circular
regression trees a more stable and regularized circular regression forest can
be built which also allows for modeling smooth effects.

@@

% -------------------------------------------------------------------
% Introduction
% -------------------------------------------------------------------
\begin{document}
\section{Introduction}
\label{sec:introduction}
Circular data can be found in a variety of applications and subject areas,
e.g., hourly crime rate in the social-economics, animal movement direction or
gene-structure in biology, and wind direction as one of the most important
weather variables in meteorology. Fitting a statistical model to this type of
data requires the incorporation of its specific feature of periodicity, i.e.,
in case of angular data the fact that values are restricted to an interval such
as $[0,2\pi)$ with $0$ being equivalent to $2\pi$.

Many approaches to model circular data assume that the circular variable of
interest follows a circular distribution, in particular the von Mises
distribution which is also known as the ``circular normal distribution''. One
of the first regression models with a circular response variable and linear
covariates was presented by \cite{Gould:1969} where the circular mean, i.e.,
the location parameter of the von Mises distribution, is predicted by a linear
combination of covariates. \cite{Johnson+Wehrly:1978} refined this idea by
plugging in a link function transforming the linear predictor to a restricted
interval of length $2\pi$. This generalized linear model (GLM)-type approach
was further extended by \cite{Fisher+Lee:1992} and subsequently by
\cite{Fisher:1993} introducing independent GLMs for either the location or
scale parameter with appropriate link functions while keeping the other
parameter constant. Additionally, they developed a combined heteroscedastic
version where the location and scale parameter are alternately reestimated
conditional on respective sets of covariates until convergence. While all of
these models are built on well elaborated theory the employed methodology has
shown to be very challenging in its application, mainly due to the complexity
encountered in optimizing the corresponding log-likelihood function which is
not globally concave. Therefore, highly informative starting values are
crucial for these circular GLMs to converge
\citep{Pewsey+Neuhaeuser+Ruxton:2013, Gill+Hangartner:2010}. In order to avoid
this strong dependence on appropriate initial values,
\cite{Mulder+Klugkist:2017} have presented a Bayesian alternative of a
homoscedastic GLM for circular data. However, apart from potential
difficulties in the optimization procedure of circular GLMs, additive effects
and their interpretation are often not intuitive for circular responses due to
the highly non-linear link function and the not-straightforward representation
of smooth transitions on the unit circle. For example, the same rotation can be
obtained in either positive or negative direction on the circle leading to an
ambiguous interpretation.

@@

This novel approach to circular regression trees and forests complements the
literature on tree-based circular modeling. \cite{Lund:2002} has already
introduced the concept of circular regression trees where binary splits are
made based on an angular distance measure employed to evaluate node
homogeneity. While this approach does not only allow for linear, but also for
circular covariates, the tree solely considers effects in the mean direction
for splitting, i.e., only the location parameter $\mu$ is regressed on
covariates. Therefore, the introduced distributional approach to circular
regression trees extends the concept of previous circular trees by considering
splits in all distribution parameters and providing a full probabilistic model.
In addition, as a natural extension, ensembles or forests of circular
regression trees are presented in this study.

\subsubsection*{Motivating example}
To provide a first impression of the presented methodology a circular
regression tree is employed for probabilistic wind direction forecasting. Wind
direction is a classical circular quantity and accurate forecasts are of great
importance for decision-making processes, e.g., in air traffic management as
considered in this study. Figure~\ref{fig:tree_ibk} shows an estimated tree
for 1-hourly forecasts at the airport of Innsbruck which lies at the bottom of
a narrow valley within the European Alps. Topography channels the wind along
the west-east valley axis or along a tributary valley intersecting from the
south. Hence, pressure gradients to which valley wind regimes react are
considered as covariates along with other meteorological measurements (lagged
by one hour) and their derivatives, such as wind direction and wind speed at
the airport or spatial and temporal differences.
%%
\begin{figure}[p!] \centering
\includegraphics[width=0.9\textheight,,angle=90,origin=c]{_plot_circforest_finalexampletree_ibk_lag1_v14.pdf}
\caption{Fitted tree based on the von Mises distribution for 1-hourly wind
direction forecasts at the airport of Innsbruck. In each terminal node the
empirical histogram (gray) and fitted density (red line) are depicted along
with the estimated location parameter (red hand). The covariates selected for
splitting are wind direction (degree), wind speed ($\text{ms}^{-1}$), and
pressure gradients ($\text{dpressure; hPa}$) west, east and south of the
airport, all lagged by one hour. Note that in the meteorological context wind
direction is defined on the scale $[0,360]$ degree and increases clockwise from
North ($0$ degree). } \label{fig:tree_ibk} \end{figure}
%%
Together with the tree structure itself, Figure~\ref{fig:tree_ibk} depicts the
empirical (gray) and fitted von Mises (red) distribution of wind direction in
each terminal node. Based on the fitted location parameters $\hat \mu$, the
subgroups can be distinguished into the following wind regimes: (1)~Up-valley
winds blowing from the valley mouth towards the upper valley (from east to
west, nodes~4 and 5); (2) Downslope winds blowing across the Alpine crest along
the intersecting valley towards Innsbruck (from south-east to north-west,
node~8); (3) Down-valley winds blowing in the direction of the valley mouth
(from west to east, nodes 10, 12 and 13). Node~7 captures observations with
rather low wind speeds that cannot be clearly distinguished into specific wind
regimes and are consequently associated with a very low estimated concentration
parameter~$\hat{\kappa}$, i.e., a high estimated variance. In terms of
covariates, the lagged wind direction (``persistence'') is mostly responsible
for distinguishing the broad range of wind regimes listed above while the
pressure gradients and wind speed separate the data into subgroups with high
vs.\ low precision. A more extensive case study of circular regression trees
and forests applied to probabilistic forecasting of wind direction at the
airports of Innsbruck and Vienna will be presented in this paper. Moreover,
their performances are benchmarked to alternative commonly used approaches.

The remainder of the paper is structured as follows: The theory on
probabilistic circular modeling introducing the von Mises distribution, and
circular regression models will be discussed in Section~\ref{sec:prob_circ}.
The methodology of circular regression trees and forests and their features are
introduced in Section~\ref{sec:tree_forest}. After the case study presented in
\ref{sec:wind}, a comprehensive summary and conclusion is given in
Section~\ref{sec:summary}.

% -------------------------------------------------------------------
% Probabilistic circular modeling
% -------------------------------------------------------------------
\section{Probabilistic circular modeling}
\label{sec:prob_circ}
Probabilistic modeling of circular data requires the selection of a proper
circular probability distribution, i.e., a probability distribution which
accounts for the periodicity of circular data. Generally, this feature can be
obtained by ``wrapping'' the probability density function of any continuous
distribution around the unit circle \citep{Mardia+Jupp:2009}. In that way, the
wrapped Cauchy distribution or the wrapped normal distribution can be employed
to model symmetric unimodal circular data. A close approximation to the wrapped
normal distribution that is mathematically simpler and hence easier to use
\citep{Fisher:1993} is provided by the von Mises distribution, a purely
circular distribution which is also known as ``the circular normal
distribution'' and is a common choice for probabilistic modeling of circular
data. Based on a location parameter $\mu \in [0, 2\pi)$ and a concentration
parameter $\kappa > 0$ the density of the von Mises distribution for an
observation $y \in [0, 2\pi)$ is given by
%%
\begin{equation}
  f_\mathrm{vM}(y; \mu, \kappa) = \frac{1}{2 \pi I_0(\kappa)}~e^{ \kappa \cos(y - \mu)},\label{schlosser:equ_vm}
\end{equation}
%%
where $I_0(\kappa)$ is the modified Bessel function of the first kind and order
$0$ (see, e.g., \citealt{Jammalamadaka+Sengupta:2001} and
\citealt{Ley+Verdebout:2017} for a more detailed overview).

@@

Once an appropriate distribution family is selected, fitting a probabilistic
model corresponds to estimating the distribution parameters. Thus, when
applying the von Mises distribution to fit a probabilistic model $\text{vM}(Y;
\mu, \kappa)$ to a circular response variable $Y \in \mathcal{Y}$ the
distribution parameters $\mu$ and $\kappa$ are to be estimated. For a learning
sample with $n$ observations $\{y_i\}_{i=1,\ldots,n}$ this can be done by
maximizing the log-likelihood function $\ell(\mu, \kappa; y)$, hence
\begin{equation}\label{eq:optim} (\hat{\mu},\hat{\kappa}) =
\argmax_{\mu,\kappa} \sum_{i=1}^n \ell(\mu, \kappa; y_i) \end{equation}
yielding maximum likelihood estimators $\hat{\mu}$ and $\hat{\kappa}$ such that
a fully specified distributional model is fitted to the learning data.

By means of the score function
%%
\begin{equation}
  \label{eq:scores}
  \begin{aligned}
    s(\mu, \kappa, y) &= \bigg(\frac{\partial \ell}{\partial \mu}(\mu,\kappa; y)
    \smallskip &,
    \quad & \frac{\partial \ell}{\partial \kappa}(\mu,\kappa; y) \bigg) 
    \\
    &= \bigg(\kappa \sin(y-\mu)
    \smallskip &,
    \quad &-\frac{I_1(\kappa)}{2\pi I_0(\kappa)}+\cos(y-\mu) \bigg)
  \end{aligned}
\end{equation}
%%
an indicator of deviation and in that way a measure of goodness of fit of the
model can be obtained for each observation and fitted parameter. Moreover, the
optimization problem in Equation~\ref{eq:optim} can also be defined by 
%%
\begin{equation}
  \sum_{i = 1}^n s(\hat{\mu}, \hat{\kappa}, y_i) = 0.
\end{equation}
%%
%%
\begin{figure}[t]
  \begin{minipage}{.05\textwidth}
    \hfill
  \end{minipage}%
  \begin{minipage}{.45\textwidth}
    \vspace{-1em}
    \includegraphics[width=1.05\linewidth]{density_linear.pdf}
  \end{minipage}%
  \begin{minipage}{.45\textwidth}
    \includegraphics[width=1.2\linewidth]{density_circular.pdf}
  \end{minipage}
  \begin{minipage}{.05\textwidth}
    \hfill
  \end{minipage}%
  \caption{Illustration of a distributional model fitted to circular data by
    maximum likelihood estimation employing the von Mises distribution. Left
    panel: linearized illustration restricted to the interval $[0, 2\pi)$. Right
    panel: circular illustration. In both panels the empirical histogram (gray)
    and fitted density (red line) are depicted along with the estimated location
    parameter (red hand).}
  \label{fig:densities} 
\end{figure}
%%

Figure~\ref{fig:densities} depicts a distributional model fitted to circular
data employing the von Mises distribution. While the left panel only serves as
an illustration of the corresponding ``linear'' description restricted to an
interval of length $2\pi$ but consequently not accounting for periodicity, the
right panel displays an appropriate representation of a circular quantity on
the unit circle. Looking at the empirical histogram along with the fitted
density (red line) and estimated location parameter (red hand) this example
shows that reasonably well fitting probabilistic models can be established for
circular data by specifying an appropriate distribution family such as the von
Mises distribution and estimating its parameters via maximum likelihood.
However, this procedure considers only the response variable. Including
covariates can improve any model remarkably as they might provide valuable
additional information.  Therefore, (distributional) regression models are of
high interest for all types of data, including circular data.

In most generalized linear or additive approaches to circular regression
modeling the location parameter $\mu$ is regressed on covariates $\mathbf{z}$
in the form of 
\begin{equation}
\mu = \mu_0 + g\left(\mathbf{\beta}^{\top} \mathbf{z}\right),
\end{equation}
with a circular intercept $\mu_0$ and a coefficient vector $\mathbf{\beta}$;
whereby the additive predictor is transformed to an interval of length $2\pi$
by a link function $g$. As an appropriate link function typically $g(x) = 2
\cdot arctan(x)$ is employed, as suggested by \citet{Fisher+Lee:1992}. While
most of the proposed models estimate a constant concentration parameter, hence,
assume homoscedasticity, \citet{Fisher+Lee:1992} also developed a
heteroscedastic version by combining two individual GLMs, each for one of the
parameters $\mu$ and $\kappa$. This provides a first approach to a fully
probabilistic regression model for circular data, however, the parameters are
not regressed simultaneously on covariates as known from the more general
framework provided by generalized additive models for location, scale, and
shape \citep[GAMLSS, ][]{Rigby+Stasinopoulos:2005}. Nevertheless, all circular
additive models incorporate the previously discussed difficulties induced by
the characteristics of the log-likelihood function and the strongly non-linear
link function. Referring to additive models in general, it has to be
considered that a proper model specification can be very challenging,
particularly for a high number of covariates and no information on possible
interactions. Moreover, the additive structure might impose a smooth effect
even if the true underlying effect is an abrupt shift.

An alternative to additive models, which can deal or even avoid the mentioned
difficulties in circular regression modeling are tree-based models. For the
particular case of circular data, the methodology of circular regression trees
and forests is presented in the following section.

@@

The main goal of employing tree-based models to circular data is to provide a
flexible and easy-to-apply modeling approach which accounts for a proper
representation on the unit circle. Based on ideas of \citet{Lund:2002}, we
introduce probabilistic circular regression trees and forests employing the von
Mises distribution. The resulting tree-based models provide a very intuitive
and data-driven alternative to commonly used regression models for circular
data.

\subsection{Circular regression trees}
\label{sec:circtree}
Fitting a global model to a full data set can be very challenging, particularly
for complex data with strong variations. Therefore, separating the data set
into more homogeneous subgroups based on covariates before fitting a local
model in each of these subgroups allows to capture (possibly) group specific
effects more precisely and hence can result in an overall better fitting model.
This is the general idea of regression trees which are combined with
distributional modeling in \cite{Schlosser+Hothorn+Stauffer:2019} based on the
unbiased recursive partitioning algorithms MOB
\citep{Zeileis+Hothorn+Hornik:2008} or CTree
\citep{Hothorn+Hornik+Zeileis:2006}. By applying the presented distributional
trees, a full distributional model is specified in each node of the tree for a
specific distribution family chosen in advance. Hence, selecting the von Mises
distribution allows for an application of distributional trees to circular data
following the tree building algorithm as introduced in Section~2.2 of
\cite{Schlosser+Hothorn+Stauffer:2019}.

In the applied tree-building algorithm, in each node the crucial decision on
how and where to split the data is based on model scores which are obtained by
evaluating the score function~$s$ for the individual observations and parameter
estimates (Equation~\ref{eq:scores}). For the von Mises distribution with its
two distribution parameters ($\mu$ and $\kappa$) and a data set of $n$
observations, this yields an $n \times 2$ matrix that can be employed as a
discrepancy measure, capturing how well each given observation conforms with
the estimated location $\hat{\mu}$ and precision $\hat{\kappa}$, respectively.
To capture dependence on covariates, the association between the model's scores
and each available covariate is assessed using either a parameter instability
test~(MOB) or a permutation test~(CTree). By doing so in each partitioning
step, the covariate with the highest significant association (i.e., lowest
significant $p$-value, if any) is selected for splitting the data. The
corresponding split point is chosen either by optimizing the
log-likelihood~(MOB) or a two-sample test statistic~(CTree) over all possible
partitions. This procedure is repeated recursively until there are no
significant parameter instabilities or until another stopping criterion is met
(e.g., subgroup size or tree depth). A more detailed description of the
applied tree-building algorithm can be found in Appendix~\ref{app:algorithm}.

Once a distributional tree model is fitted it can be employed to obtain
probabilistic predictions for a possibly new set of observed covariates $\bm{z}
= (z_1, \ldots, z_m)$. Starting at the root node, the tree structure leads the
observation to a terminal node, depending on the values of the covariates,
where the parameter pair $(\hat{\mu}, \hat{\kappa})$ is estimated for the
corresponding subset of learning observations. This can also be expressed by
employing weights which indicate whether the $i$-th learning observation and
the observation $\bm{z}$ belong to the same terminal node: \begin{equation}
w^{\text{tree}}_i(\bm{z}) = \sum_{b=1}^B \mathbf{1}((\bm{z}_i \in
\mathcal{B}_b) \land (\bm{z} \in \mathcal{B}_b)), \end{equation} where
$\mathbf{1}(\cdot)$ is the indicator function and $\mathcal{B}_b$ is the $b$-th
out of $B$ segments partitioning the covariate space in disjoint subsets. In
that way the estimated parameter pair $(\hat{\mu},\hat{\kappa})(\bm{z})$
specifying the predicted von Mises distribution for a given $\bm{z}$ is
obtained by 
%%
\begin{equation} 
  (\hat{\mu},\hat{\kappa})(\bm{z}) =
  \argmax_{\mu,\kappa} \sum_{i=1}^n w^{\text{tree}}_i(\bm{z}) \cdot
  \ell(\mu,\kappa; y_i).
\end{equation}
%%
Therefore, the same parameter pair is estimated for all observations belonging
to the same terminal node. This facilitates the application as the parameter
estimates do not need to be recalculated for each (new) observation via maximum
likelihood but can be extracted directly from the learning sample and the
fitted model.

While tree models can capture non-additive effects, their structure and the
consequential strict separation of data into subgroups hinders an adequate
depiction of smooth effects. Therefore, they might impose abrupt changes even
in case of rather smooth transitions. This can be avoided by combining an
ensemble of trees in order to obtain a more stable forest model.

>>>>

\Abstract{
While circular data occur in a wide range of scientific fields,
the methodology for probabilistically forecasting them is
highly limited. Most of the existing methods are built on the framework of
generalized linear and additive models, which can incorporate difficulties in
both the optimization and the interpretation. As a very intuitive alternative
to additive circular regression models, we extend previous ideas of circular
trees and introduce probabilistic circular regression trees and forests
employing the von Mises distribution. The resulting tree-based models simplify
the estimation process by separating the data into more homogeneous subgroups
and fitting a simple intercept-only model for each resulting subset. By
incorporating the framework of distributional modeling, splits in all
distribution parameters are considered and a fully specified distribution is
provided for each terminal node. These circular regression trees are
easy to interpret, can detect non-additive effects, and select covariates and
their interactions automatically. By combining an ensemble of circular
regression trees a more stable and regularized circular regression forest can
be built, which also allows for modeling smooth effects.

@@

% -------------------------------------------------------------------
% Introduction
% -------------------------------------------------------------------
\begin{document}
\section{Introduction}
\label{sec:introduction}
Circular data can be found in a variety of applications and subject areas,
e.g., hourly crime rate in the social-economics, animal movement direction or
gene-structure in biology, and wind direction as one of the most important
weather variables in meteorology. Fitting a statistical model to this type of
data requires the incorporation of its specific feature of periodicity. For example,
angular data are restricted to an interval such
as $[0,2\pi)$ with $0$ being equivalent to $2\pi$.

Many approaches to model circular data assume that the circular variable of
interest follows a circular distribution, in particular the von Mises
distribution which is also known as the ``circular normal distribution''. One
of the first regression models with a circular response variable and linear
covariates was presented by \cite{Gould:1969} where the circular mean, i.e.,
the location parameter of the von Mises distribution, is predicted by a linear
combination of covariates. \cite{Johnson+Wehrly:1978} refined this idea by
plugging in a link function transforming the linear predictor to a restricted
interval of length $2\pi$. This generalized linear model (GLM)-type approach
was further extended by \cite{Fisher+Lee:1992} and subsequently by
\cite{Fisher:1993} introducing independent GLMs for either the location or
scale parameter with appropriate link functions while keeping the other
parameter constant. Additionally, they developed a combined heteroscedastic
version where the location and scale parameter are alternately reestimated
conditional on respective sets of covariates until convergence. While all of
these models are built on well elaborated theory the employed methodology has
shown to be very challenging in its application, mainly due to the complexity
encountered in optimizing the corresponding log-likelihood function which is
not globally concave. Therefore, highly informative starting values are
crucial for these circular GLMs to converge
\citep{Pewsey+Neuhaeuser+Ruxton:2013, Gill+Hangartner:2010}. In order to avoid
this strong dependence on appropriate initial values,
\cite{Mulder+Klugkist:2017} presented a Bayesian alternative of a
homoscedastic GLM for circular data. However, apart from potential
difficulties in the optimization procedure of circular GLMs, additive effects
and their interpretation are often not intuitive for circular responses due to
the highly non-linear link function and the not-straightforward representation
of smooth transitions on the unit circle. For example, the same rotation can be
obtained in either positive or negative direction on the circle leading to an
ambiguous interpretation.

@@

This novel approach to circular regression trees and forests complements the
literature on tree-based circular modeling. \cite{Lund:2002} already
introduced the concept of circular regression trees where binary splits are
made based on an angular distance measure employed to evaluate node
homogeneity. While this approach does not only allow for linear, but also for
circular covariates, the tree solely considers effects in the mean direction
for splitting so that only the location parameter $\mu$ is regressed on
covariates. Therefore, the introduced distributional approach to circular
regression trees extends the concept of previous circular trees by considering
splits in all distribution parameters and providing a full probabilistic model.
In addition, as a natural extension, ensembles or forests of circular
regression trees are presented in this study.

\subsubsection*{Motivating example}
To provide a first impression of the presented methodology, a circular
regression tree is employed for probabilistic wind direction forecasting. Wind
direction is a classical circular quantity and accurate forecasts are of great
importance for decision-making processes, e.g., in air traffic management as
considered in this study. Figure~\ref{fig:tree_ibk} shows an estimated tree
for 1-hourly forecasts at the airport of Innsbruck which lies at the bottom of
a narrow valley within the European Alps. Topography channels the wind along
the west-east valley axis or along a tributary valley intersecting from the
south. Hence, pressure gradients to which valley wind regimes react are
considered as covariates along with other meteorological measurements (lagged
by one hour) and their derivatives, such as wind direction and wind speed at
the airport or spatial and temporal differences.
%%
\begin{figure}[p!] \centering
\includegraphics[width=0.9\textheight,,angle=90,origin=c]{_plot_circforest_finalexampletree_ibk_lag1_v14.pdf}
\caption{Fitted tree based on the von Mises distribution for 1-hourly wind
direction forecasts at the airport of Innsbruck. In each terminal node the
empirical histogram (gray) and fitted density (red line) are depicted along
with the estimated location parameter (red hand). The covariates selected for
splitting are wind direction (degree), wind speed ($\text{ms}^{-1}$), and
pressure gradients ($\text{dpressure; hPa}$) west, east and south of the
airport, all lagged by one hour. Note that in the meteorological context wind
direction is defined on the scale $[0,360]$ degree and increases clockwise from
North ($0$ degree). } \label{fig:tree_ibk} \end{figure}
%%
Figure~\ref{fig:tree_ibk} illustrates the obtained tree structure together with
the empirical (gray) and fitted von Mises (red) distribution of wind direction
in each terminal node. Based on the fitted location parameters $\hat \mu$, the
subgroups can be distinguished into the following wind regimes: (1)~Up-valley
winds blowing from the valley mouth towards the upper valley (from east to
west, nodes~4 and 5); (2) Downslope winds blowing across the Alpine crest along
the intersecting valley towards Innsbruck (from south-east to north-west,
node~8); (3) Down-valley winds blowing in the direction of the valley mouth
(from west to east, nodes 10, 12 and 13). Node~7 captures observations with
rather low wind speeds that cannot be clearly distinguished into specific wind
regimes and are consequently associated with a very low estimated concentration
parameter~$\hat{\kappa}$, i.e., a high estimated variance. In terms of
covariates, the lagged wind direction (``persistence'') is mostly responsible
for distinguishing the broad range of wind regimes listed above while the
pressure gradients and wind speed separate the data into subgroups with high
vs.\ low precision. A more extensive case study of circular regression trees
and forests applied to probabilistic forecasting of wind direction at the
airports of Innsbruck and Vienna will be presented in Section~\ref{sec:wind} of
the paper. Moreover, their performances are benchmarked to alternative commonly
used approaches.

The remainder of the paper is structured as follows: The theory on
probabilistic circular modeling introducing the von Mises distribution, and
circular regression models will be discussed in Section~\ref{sec:prob_circ}.
The methodology of circular regression trees and forests and their features are
introduced in Section~\ref{sec:tree_forest}. After the case study presented in
Section~\ref{sec:wind}, a comprehensive summary and conclusions are given in
Section~\ref{sec:summary}.

% -------------------------------------------------------------------
% Probabilistic circular modeling
% -------------------------------------------------------------------
\section{Probabilistic circular modeling}
\label{sec:prob_circ}
Probabilistic modeling of circular data requires the selection of a probability
distribution which accounts for the periodicity of circular data. Generally,
this feature can be obtained by ``wrapping'' the probability density function
of any continuous distribution around the unit circle \citep{Mardia+Jupp:2009}.
In that way, the wrapped Cauchy distribution or the wrapped normal distribution
can be employed to model symmetric unimodal circular data. A close
approximation to the wrapped normal distribution that is mathematically simpler
and hence easier to use \citep{Fisher:1993} is provided by the von Mises
distribution, a purely circular distribution which is also known as ``the
circular normal distribution'' and is a common choice for probabilistic
modeling of circular data. Based on a location parameter $\mu \in [0, 2\pi)$
and a concentration parameter $\kappa > 0$ the density of the von Mises
distribution for an observation $y \in [0, 2\pi)$ is given by
%%
\begin{equation}
  f_\mathrm{vM}(y; \mu, \kappa) = \frac{1}{2 \pi I_0(\kappa)}~e^{ \kappa \cos(y - \mu)},\label{schlosser:equ_vm}
\end{equation}
%%
where $I_0(\kappa)$ is the modified Bessel function of the first kind and order
$0$ (see, e.g., \citealt{Jammalamadaka+Sengupta:2001} and
\citealt{Ley+Verdebout:2017} for a more detailed overview).

@@

To fit a probabilistic model $\text{vM}(Y;
\mu, \kappa)$ to a circular response variable $Y \in \mathcal{Y}$, the
distribution parameters $\mu$ and $\kappa$ are to be estimated.
This can be done by
maximizing the log-likelihood function $\ell(\mu, \kappa; y)$
%%
\begin{equation}\label{eq:optim}
  (\hat{\mu},\hat{\kappa}) = \argmax_{\mu,\kappa} \sum_{i=1}^n \ell(\mu, \kappa; y_i) 
\end{equation}
%%
yielding maximum likelihood estimators $\hat{\mu}$ and $\hat{\kappa}$ such that
a fully specified distributional model is fitted to the learning data $\{y_i\}_{i=1,\ldots,n}$.

The score function
%%
\begin{equation}
  \label{eq:scores}
  \begin{aligned}
    s(\mu, \kappa, y) &= \bigg(\frac{\partial \ell}{\partial \mu}(\mu,\kappa; y)
    \smallskip &,
    \quad & \frac{\partial \ell}{\partial \kappa}(\mu,\kappa; y) \bigg) 
    \\
    &= \bigg(\kappa \sin(y-\mu)
    \smallskip &,
    \quad &-\frac{I_1(\kappa)}{2\pi I_0(\kappa)}+\cos(y-\mu) \bigg)
  \end{aligned}
\end{equation}
%%
provides a way to obtain a measure of goodness of fit of the model for
each observation and fitted parameter. Then, the optimization problem in
Equation~\ref{eq:optim} can alternatively be specified as 
%%
\begin{equation}
  \sum_{i = 1}^n s(\hat{\mu}, \hat{\kappa}, y_i) = 0.
\end{equation}
%%
%%
\begin{figure}[t]
  \begin{minipage}{.05\textwidth}
    \hfill
  \end{minipage}%
  \begin{minipage}{.45\textwidth}
    \vspace{-1em}
    \includegraphics[width=1.05\linewidth]{density_linear.pdf}
  \end{minipage}%
  \begin{minipage}{.45\textwidth}
    \includegraphics[width=1.2\linewidth]{density_circular.pdf}
  \end{minipage}
  \begin{minipage}{.05\textwidth}
    \hfill
  \end{minipage}%
  \caption{Illustration of a distributional model fitted to circular data by
    maximum likelihood estimation employing the von Mises distribution. Left
    panel: linearized illustration restricted to the interval $[0, 2\pi)$. Right
    panel: circular illustration. In both panels the empirical histogram (gray)
    and fitted density (red line) are depicted along with the estimated location
    parameter (red hand).}
  \label{fig:densities} 
\end{figure}
%%

Figure~\ref{fig:densities} depicts two representations of the same
distributional model fitted to circular data employing the von Mises
distribution: While the left panel only serves as an illustration of the
corresponding ``linear'' description restricted to an interval of length $2\pi$
but consequently not accounting for periodicity, the right panel displays an
appropriate representation of a circular quantity on the unit circle. Looking
at the empirical histogram along with the fitted density (red line) and
estimated location parameter (red hand) this example shows that reasonably well
fitting probabilistic models can be established for circular data by specifying
an appropriate distribution family such as the von Mises distribution and
estimating its parameters via maximum likelihood.  However, this procedure
considers only the response variable. Including covariates can improve any
model remarkably as they might provide valuable additional information.
Therefore, (distributional) regression models are of high interest for all
types of data, including circular data.

In most generalized linear or additive approaches to circular regression
modeling the location parameter $\mu$ is regressed on covariates $\mathbf{z}$
in the form of 
\begin{equation}
\mu = \mu_0 + g\left(\mathbf{\beta}^{\top} \mathbf{z}\right),
\end{equation}
with a circular intercept $\mu_0$ and a coefficient vector $\mathbf{\beta}$.  A
link function $g$ transforms the additive predictor to an interval of length
$2\pi$.  Typically $g(x) = 2\cdot arctan(x)$ is employed, as suggested by
\citet{Fisher+Lee:1992}.  They also developed a heteroscedastic version by
combining two individual GLMs, each for one of the parameters $\mu$ and
$\kappa$. This provides a first approach to a fully probabilistic regression
model for circular data. However, the parameters are not regressed
simultaneously on covariates as in the more general framework provided by
generalized additive models for location, scale, and shape \citep[GAMLSS,
][]{Rigby+Stasinopoulos:2005}. Nevertheless, all circular additive models
incorporate the previously discussed difficulties induced by the
characteristics of the log-likelihood function and the strongly non-linear link
function. Referring to additive models in general, it has to be considered that
a proper model specification can be very challenging, particularly for a high
number of covariates and no information on possible interactions. Moreover, the
additive structure might impose a smooth effect even if the true underlying
effect is an abrupt shift, which occurs, e.g., in atmospheric wind fields.

An alternative to additive models, which can deal or even avoid the mentioned
difficulties in circular regression modeling are tree-based models.

@@

Based on ideas of \citet{Lund:2002}, we introduce circular regression trees and
forests considering splits in all distribution parameters of the von Mises
distribution and providing a full probabilistic model.  Moreover, the resulting
tree-based models provide a very intuitive and data-driven alternative to
commonly used GLMs for circular data.

\subsection{Circular regression trees}
\label{sec:circtree}
Fitting a global model to a full data set can be very challenging, particularly
for complex data with strong variations. Therefore, separating the data set
into more homogeneous subgroups based on covariates before fitting a local
model in each of these subgroups allows to capture (possibly) group specific
effects more precisely and hence can result in an overall better fitting model.
This is the general idea of regression trees which are combined with
distributional modeling in \cite{Schlosser+Hothorn+Stauffer:2019}. Specifying a
full distributional model in each node of the tree yields a distributional
regression tree, where selecting the von Mises distribution enables an
application to circular data. The crucial step of how and where to split the
data can be accomplished with the unbiased recursive partitioning algorithms
MOB \citep{Zeileis+Hothorn+Hornik:2008} or CTree
\citep{Hothorn+Hornik+Zeileis:2006}. For this purpose, model scores are
obtained by evaluating the score function~$s$ for the individual observations
and parameter estimates (Equation~\ref{eq:scores}). For the von Mises
distribution with its two distribution parameters ($\mu$ and $\kappa$) and a
data set of $n$ observations, this yields an $n \times 2$ matrix that can be
employed as a discrepancy measure, capturing how well each given observation
conforms with the estimated location $\hat{\mu}$ and precision $\hat{\kappa}$,
respectively.  To capture dependence on covariates, the association between the
model's scores and each available covariate is assessed using either a
parameter instability test~(MOB) or a permutation test~(CTree). By doing so in
each partitioning step, the covariate with the highest significant association
(i.e., lowest significant $p$-value, if any) is selected for splitting the
data. The corresponding split point is chosen either by optimizing the
log-likelihood~(MOB) or a two-sample test statistic~(CTree) over all possible
partitions. This procedure is repeated recursively until there are no
significant parameter instabilities or until another stopping criterion is met
(e.g., subgroup size or tree depth). A more detailed description of the applied
tree-building algorithm can be found in Appendix~\ref{app:algorithm}.

Once a distributional tree model is fitted it can be employed to obtain
probabilistic predictions for a possibly new set of observed covariates $\bm{z}
= (z_1, \ldots, z_m)$. Starting at the root node, the tree structure leads the
observation to a terminal node where the parameter pair $(\hat{\mu},
\hat{\kappa})$ is estimated for the corresponding subset of learning
observations. This can also be expressed by employing weights which indicate
whether the $i$-th learning observation and the observation $\bm{z}$ belong to
the same terminal node: 
%%
\begin{equation}
  w^{\text{tree}}_i(\bm{z}) = \sum_{b=1}^B \mathbf{1}((\bm{z}_i \in
  \mathcal{B}_b) \land (\bm{z} \in \mathcal{B}_b)), 
\end{equation} 
%%
where $\mathbf{1}(\cdot)$ is the indicator function and $\mathcal{B}_b$ is the
$b$-th out of $B$ segments partitioning the covariate space in disjoint
subsets. Then the estimated parameter pair
$(\hat{\mu},\hat{\kappa})(\bm{z})$ specifying the predicted von Mises
distribution for a given $\bm{z}$ is obtained by 
%%
\begin{equation} 
  (\hat{\mu},\hat{\kappa})(\bm{z}) =
  \argmax_{\mu,\kappa} \sum_{i=1}^n w^{\text{tree}}_i(\bm{z}) \cdot
  \ell(\mu,\kappa; y_i).
\end{equation}
%%
Therefore, the same parameter pair is estimated for all observations belonging
to the same terminal node, which speeds up computation since the parameter
estimates do not need to be recalculated for each (new) observation via maximum
likelihood but can be extracted directly from the learning sample and the
fitted model.

While tree models can capture non-additive effects, their structure and the
consequential strict separation of data into subgroups hinders an adequate
depiction of smooth effects. They can be included by combining an ensemble of
trees in order to obtain a regression forest, which also stabilizes the model.
